
# Determining Corelation Between Food Price Inflation and Population Growth Over the Decades

```{r}
# Install packages if not already installed
if (!require(readxl)) install.packages('readxl')
if (!require(tidyverse)) install.packages('tidyverse')
if (!require(dplyr)) install.packages('dplyr')
if (!require(zoo)) install.packages('zoo')
if (!require(gplots)) install.packages('gplots')
if (!require(heatmaply)) install.packages('heatmaply')


# Load packages
library(readxl)
library(tidyverse)
library(dplyr)
library(zoo)
library(gplots)
library(heatmaply)
```


#Step 1.Loading and Preprocessing the Dataset

#Step 1.1 Loading the Dataset

```{r}
#loading Food price index
food_index <- read_excel("F:/DMTT/A1/FoodPrices_2013 to 2023.xlsx", col_names = TRUE)
```


```{r}
#printing food_index data
print(food_index)
```


```{r}
#loading U.S. census data
population_data <- read_excel("F:/DMTT/A1/US Population Data 1790 to 2020.xlsx", col_names = c('decade', 'Total_Population', 'population_increase', 'population_increase_%', 'Urban_%', 'Rural_%'))
```


```{r}
#print population_data
print(population_data)
```


```{r}
# remove the first and second rows
population_data <- population_data[-c(1, 2), ]
print(population_data)
```


#Step 1.2 Preprocessing the data

#Step 1.21 Checking for specical characters

```{r}
special_char <- function(df, col) {
    special_cha <- grepl("[!@#$^&*()%]", df[[col]])
    if (any(special_cha)) {
        cat(paste("Special characters found in column '", col, "':\n"))
        print(df[special_cha, ])
    }
}
```


```{r}
# apply the function to all columns in your dataframe
lapply(names(population_data), function(col) special_char(population_data, col))

```


```{r}
#checking in food_index
lapply(names(food_index), function(col) special_char(food_index, col))
```

#Step 1.22 Removing special characters

```{r}
# function to remove special characters
remove_special_chars <- function(df){
  df[] <- lapply(df, function(x) gsub('[!@#$%^&*()]', '', x))
  return(df)
}
```


```{r}
# removing special characters from food_index
food_index <- remove_special_chars(food_index)
```


```{r}
# checking for special characters in food_index
lapply(names(food_index), special_char, df = food_index)
```


```{r}
# displaying food_index
tail(food_index)
```

```{r}
# replacing value for inflation rate at index 1 with 0 (R uses 1-based indexing)
food_index$`Inflation Rate`[1] <- 0
head(food_index)
```


#Step 1.23 checking for missing value and replacing missing value

```{r}
# checking for null value in food_index
original_colnames <- colnames(food_index)
food_index <- data.frame(sapply(food_index, function(x) ifelse(x == 'NA', NA, x)))
colnames(food_index) <- original_colnames
colSums(is.na(food_index))
```


```{r}
# checking for null value in population_data
original_colnames <- colnames(population_data)
population_data <- data.frame(sapply(population_data, function(x) ifelse(x == 'NA', NA, x)))
colnames(population_data) <- original_colnames
colSums(is.na(population_data))

```


```{r}
# replacing the null value through forward filling - using previous decades values
population_data$`Urban_%` <- na.locf(population_data$`Urban_%`)
population_data$`Rural_%` <- na.locf(population_data$`Rural_%`)
```


```{r}
# checking for null values
colSums(is.na(population_data))
```

#Step 1.24: Adjusting temporal scale:


```{r}
# adding 3 years (1910, 1911, 1912) to the food_index with feature values as zero
new_rows <- data.frame(
  'Year' = c(1910, 1911, 1912, 2024, 2025, 2026, 2027, 2028, 2029),
  'USD Value' = c(0,0,0,0,0,0,0,0,0),
  'Inflation Rate' = c(0,0,0,0,0,0,0,0,0)
)
names(new_rows) <- c("Year", "USD Value", "Inflation Rate")

# concatenate the new rows with the original DataFrame
food_index <- rbind(new_rows, food_index)
print(food_index)
```

```{r}
# create a new DataFrame for the decade data
decade_food_index <- data.frame()

# convert 'Year' column to integers and create a 'decade' column
food_index$Year <- as.integer(food_index$Year)
food_index$decade <- (food_index$Year %/% 10) * 10

# convert 'Inflation_Rate' column to numeric
food_index$`Inflation Rate` <- as.numeric(as.character(food_index$`Inflation Rate`))

# replacing the index 120 from which the special character is removed with correct value after converting into numeric
food_index[120, 'Inflation Rate'] <- 0.0496

# Group by 'decade' and calculate the sum of 'Inflation Rate'
decade_food_index <- food_index %>% group_by(decade) %>% summarise(`inflation_rate_%` = sum(`Inflation Rate`)*100, .groups="keep")
decade_food_index$`inflation_change_%` <- c(0, diff(decade_food_index$`inflation_rate_%`))
decade_food_index[is.na(decade_food_index)] <- 0
print(decade_food_index)
```


#Step 1.25: Merging data


```{r}
# merging data
merged_df <- merge(decade_food_index, population_data, by="decade")
merged_df$`inflation_rate_cumsum_%` <- cumsum(merged_df$`inflation_rate_%`)
merged_df$`population_increase_cumsum_%` <- cumsum(merged_df$`population_increase_%`)
print(merged_df)
```


#Step 2: Detemining if rate of increase in Food Price Inflation and Population growth over the decades are correlated


```{r}
# convert columns to numeric
merged_df$`inflation_rate_%` <- as.numeric(as.character(merged_df$`inflation_rate_%`))
merged_df$`population_increase_%` <- as.numeric(as.character(merged_df$`population_increase_%`))

# plotting
ggplot(data = merged_df) +
  geom_line(aes(x = decade, y = `inflation_rate_%`, colour = "Inflation Rate"), show.legend = TRUE) +
  geom_line(aes(x = decade, y = `population_increase_%`, colour = "Population Increase"), show.legend = TRUE) +
  labs(title = "Inflation Rate vs Population Increase % over the decades", x = "Decade", y = "Percentage %", colour = "Parameter") +
  scale_colour_manual(values = c("Inflation Rate" = "red", "Population Increase" = "blue")) +
  theme_minimal()

```


```{r}
# visualizing the cumulative foot price inflation % and population increase % over the decades to understand more about the correlation between two parameters

# convert columns to numeric
merged_df$`inflation_rate_cumsum_%` <- as.numeric(as.character(merged_df$`inflation_rate_cumsum_%`))
merged_df$`population_increase_cumsum_%` <- as.numeric(as.character(merged_df$`population_increase_cumsum_%`))

# plotting
ggplot(data = merged_df) +
  geom_line(aes(x = decade, y = `inflation_rate_cumsum_%`, colour = "Food Price Inflation Rate %"), show.legend = TRUE) +
  geom_line(aes(x = decade, y = `population_increase_cumsum_%`, colour = "Increase in U.S. Population%"), show.legend = TRUE) +
  labs(title = "Rise in U.S. Food Inflation and Population Over the Decades", x = "Decade", y = "Percentage %", colour = "Parameter") +
  scale_colour_manual(values = c("Food Price Inflation Rate %" = "red", "Increase in U.S. Population%" = "blue")) +
  theme_minimal()

```


#Step 2.1: Calculating and displaying correlation matrix

#Visualizing corelation matrix for all cumulative values


```{r}
# create a new data frame for correlation analysis
corr_df <- merged_df[, c('decade', 'inflation_rate_cumsum_%', 'population_increase_cumsum_%'), drop = FALSE]
colnames(corr_df) <- make.names(c('decade', 'inflation_rate_cumsum_percent', 'population_increase_cumsum_percent'))
# convert non-numeric columns to numeric type
corr_df <- as.data.frame(lapply(corr_df, function(x) ifelse(is.na(as.numeric(x)), x, as.numeric(x))))

# calculate the correlation matrix
corr_matrix <- cor(corr_df, method = 'pearson')
corr_matrix
```


```{r}
# visualizing correlation matrix as a heatmap
heatmaply(corr_matrix,
          col = colorRampPalette(c("blue", "white", "red"))(100),
          main = "Pearson Correlation Heatmap")
```



#Calcualting spearman correlation matrix since it is less sensitive to outliers


```{r}
sp_corr_matrix <- cor(corr_df, method = "spearman")
sp_corr_matrix
```
```{r}
# visualizing correlation matrix as a heatmap
heatmaply(sp_corr_matrix,
          col = colorRampPalette(c("blue", "white", "red"))(100),
          main = "Spearman Correlation Heatmap")
```


#Step 2.2: Visualizing scatter plot between Cumulative Inflation Rate over the decade (%) and Cumulative Population Increase over the decade (%)


```{r}
# convert the data types to numeric
merged_df$`inflation_rate_cumsum_%` <- as.numeric(as.character(merged_df$`inflation_rate_cumsum_%`))
merged_df$`population_increase_cumsum_%` <- as.numeric(as.character(merged_df$`population_increase_cumsum_%`))

# create scatter plot
plot(merged_df$`inflation_rate_cumsum_%`, merged_df$`population_increase_cumsum_%`, 
     xlab = "Cumulative Inflation Rate over the decade (%)",
     ylab = "Cumulative Population Increase over the decade (%)",
     main = "Scatter Plot: Inflation Rate vs Population Increase")

# calculate trendline using lm()
fit <- lm(merged_df$`population_increase_cumsum_%` ~ merged_df$`inflation_rate_cumsum_%`)
abline(fit, col = "red", lty = "dashed")

# adding labels and title
legend("topright", legend = "Trendline", col = "red", lty = "dashed")
```

```{r}
```

Conclusion

*Inflation Rate Cumulative Sum and Population Increase Cumulative Sum:
  - There is a very strong positive correlation between 'inflation_rate_cumsum_%' and 'population_increase_cumsum_%'.
  - This implies that as decade over decade population increases cumulative inflation also increases.
  - From the correlation matrix we could conclude that there is a strong positive relationships between these two variables.

*Inflation Rate Cumulative Sum and Decade:
  - There is a very strong positive correlation between 'inflation_rate_cumsum_%' and 'decade'.
  - This implies that inflation tends to increase as decades progresses.

*Population Increase Cumulative Sum and Decade:
  - There is a very strong positive correlation between 'population_increase_cumsum_%' and 'decade'.
  - This suggests that population also tends to increase as decades progresses.

However, further study and considering other factors influencing both parameters is required to understand the relationship between these variables as correlation does not implies causation.

